\documentclass[11pt,a4paper]{article}

% Packages
\usepackage[utf8]{inputenc}
\usepackage[T1]{fontenc}
\usepackage{lmodern}
\usepackage{amsmath,amssymb,amsthm}
\usepackage{graphicx}
\usepackage{xcolor}
\usepackage{hyperref}
\usepackage{natbib}
\usepackage{booktabs}
\usepackage{array}
\usepackage{multirow}
\usepackage{caption}
\usepackage{subcaption}
\usepackage{tikz}
\usetikzlibrary{shapes,arrows,positioning,calc,decorations.pathreplacing}
\usepackage{algorithm}
\usepackage{algorithmic}
\usepackage{enumitem}
\usepackage{geometry}
\geometry{margin=1in}
\usepackage{adjustbox}

% Bibliography style
\bibliographystyle{unsrtnat}

% Title information
\title{\textbf{From Prediction to Design: A Paradigm Shift in AI-Driven Protein-Protein Interaction Research}}
\author{
\textbf{Author Names}$^{1,*}$\\
\\
$^1$Affiliation\\
$^*$Corresponding author
}
\date{}

\begin{document}

\maketitle

\begin{abstract}
\noindent
Protein-protein interactions (PPIs) form the molecular foundation of cellular life. Over the past two decades, computational methods for PPI analysis have primarily focused on \textit{prediction}---inferring interaction properties from protein information. However, a fundamental paradigm shift is underway: from ``predicting what will happen'' to ``designing what we want.'' This transformation is driven by breakthroughs in protein language models, diffusion models, and equivariant neural networks. In this review, we argue that this transition constitutes a genuine paradigm shift in the Kuhnian sense, enabling investigation of sequence-structure space that evolution never explored. We systematically analyze the technical foundations, core design tasks, and challenges ahead. We address critical issues including evaluation bias, the hallucination problem, and dual-use considerations. We conclude with grand challenges for the field: mapping designability boundaries, designing dynamic interactions, and developing interpretable methods for clinical translation. This paradigm shift marks a fundamental change in how AI interfaces with biological discovery---from passive observation to active creation.
\end{abstract}

\noindent\textbf{Keywords:} protein-protein interactions, deep learning, generative models, protein design, AlphaFold, diffusion models, protein language models, paradigm shift

\vspace{1em}

% Table of Contents (for draft)
\tableofcontents
\newpage

% Main content sections (revised structure)
\input{sections_v2/figures}
\input{sections_v2/section1_introduction}
\input{sections_v2/section2_paradigm_transition}
\input{sections_v2/section3_design_tasks}
\input{sections_v2/section4_technologies}
\input{sections_v2/section5_challenges}
\input{sections_v2/section6_applications}
\input{sections_v2/section7_future}
\input{sections_v2/section8_conclusion}

% Bibliography
\bibliography{references}

\end{document}
